The Bot on Speaking Terms: The Effects of Conversation Architecture on Perceptions of Conversational Agents

### 2023-02-27 figures/fig-heatmap_identified.tex
submitted version to CUI 2023

\begin{figure*}[]
  \centering
  \includegraphics[width=\textwidth]{figures/fig-heatmap_identified.pdf}
  \caption{Synthesized framework on the identified relationship between perception of agents and conversation architecture elements}
  \Description{This figure shows a heatmap the 183 identified relationships between conversation architecture elements and perceptions of agents. Each speech element is listed down the rows of the heatmap, grouped by the four categories dialog strategy, content affectiveness, content style, and speech format. Each perception aspect is listed across the columns of the table, grouped by the four categories perception of interaction with agent, perception of agent's ability, perception of sociability with agent, and perception of agent's humanness. Each cell has a different shade of colour, with the darker shades depicting higher number of relationships, and the lighter shades depicting smaller number of relationships.}
  \label{fig:heatmap_identified}
\end{figure*}

>>>>

\begin{figure*}[t]
  \centering
  \includegraphics[width=0.9\textwidth]{figures/fig-heatmap_identified.pdf}
  \caption{Synthesized framework on the identified relationship between perception of agents and conversation architecture elements. The heatmap visualizes the number of relationships identified from the literature. Note that a single paper may describe more than one relationship.}
  \Description{This figure shows a heatmap the 183 identified relationships between conversation architecture elements and perceptions of agents. Each speech element is listed down the rows of the heatmap, grouped by the four categories dialog strategy, content affectiveness, content style, and speech format. Each perception aspect is listed across the columns of the table, grouped by the four categories perception of interaction with agent, perception of agent's ability, perception of sociability with agent, and perception of agent's humanness. Each cell has a different shade of colour, with the darker shades depicting higher number of relationships, and the lighter shades depicting smaller number of relationships.}
  \label{fig:heatmap_identified}
\end{figure*}

### 2023-02-27 sections/1-introduction.tex
submitted version to CUI 2023

Conversational agents (CAs), such as Amazon Alexa or Google Assistant, are designed to interact with humans using natural language through voice or text modalities. Since these agents communicate in natural language, \textit{conversation architecture}---what and how a CA says---affects users' perceptions of agents and, in turn, the effectiveness of their interactions with users \cite{knijnenburg2016inferring, seeger2021chatbots}, i.e. supporting users in achieving users' goals. For instance, previous research suggests that there is a relationship between CAs' conversation architecture and how users perceive these agents %.
%While various studies investigated how some elements of conversation architecture affect users' perceptions of agents~
(e.g.,~\cite{clark2019state, feine2019taxonomy, finch2020towards}).  However, there is currently a lack of systematic understanding of these relationships, i.e. what elements of conversation architecture are impactful for users' perceptions and how. Specifically, while various studies investigated how some elements of conversation architecture affect some aspects of users' perceptions or discussed the speech specifics of agents as part of a broader analysis of the state of research \cite{clark2019state,feine2019taxonomy}, there is no comprehensive analysis of %research specifically for 
conversation architecture elements that are relevant to user perceptions. Yet, this systematic understanding is essential for guiding the design of conversation architecture, which would allow us to both predict and develop users' appropriate perceptions of agents.

@@

%These research gaps related to the paucity of knowledge on synthesizing the effect of speech specifics on perceptions of CAs as well as the lack of systematic understanding on the elements of conversation architecture and perceptions of agents 
To address this gap, in this paper, we
%we aim to understand how to design conversation architecture in order to orchestrate perceptions of conversational agents.
%First, we 
explore the following research questions: \textbf{(RQ1) what aspects of users' perceptions of agents have been explored in relation to the effects of conversation architecture elements}, \textbf{(RQ2) what elements of conversation architecture are relevant to users' perceptions of agents?}, and \textbf{(RQ3) what is known of the specifics of these relationships between conversation architecture elements and the perceptions of CAs?}
%While there are various elements of conversation architecture used in a CA's utterances, we don't have a systematic understanding of them. This leads to our second research question: \textbf{(RQ2) what elements of conversation architecture are relevant to users' perceptions of agents?} Lastly, we want to understand the effect of conversation architecture on perceptions of agent, specifically: \textbf{(RQ3) what are the relationships between conversation architecture elements and the perceptions of CAs?}
To answer these questions, we performed a comprehensive review of the existing literature published from 2010 to 2022 and synthesized the effect of conversation architecture elements on the perceptions of CAs based on the final dataset of 57 relevant papers (\autoref{fig:heatmap_identified}). Based on this review, we grouped the perceptions of agents that were shown to be affected by conversation architecture into four categories and eleven aspects (\autoref{tbl:perceptions}) and grouped the speech variations related to perceptions of agents into four categories and eleven elements (\autoref{tbl:conversation_architecture}). 
We found that perceptions of interaction with CAs (usability, engagement, satisfaction) have been most explored in relation to speech variations, while the perceptions of CAs' ability (intelligence, competence, credibility) have been explored the least.
%the least number of explored and identified relationships. 
Our results also show that the conversation architecture elements of response delay and agent-initiated dialog-related perceptions of CAs are under-explored in literature. 
Thus, this work contributes to the CUI community by presenting a framework that synthesizes the findings on the effect of conversation architecture on perceptions of conversational agents and identifies and clusters perception aspects and conversation architecture elements. %, providing clear and consistent terms that can be used in future research. Together, these contributions move us one step closer to designing CAs that effectively interact with users.

>>>>

Conversational agents (CAs), such as Amazon Alexa or Google Assistant, are designed to interact with humans using natural language through voice or text modalities. Since these agents communicate in natural language, \textit{conversation architecture}---what and how a CA says---affects users' perceptions of agents and, in turn, the effectiveness of their interactions with users \cite{knijnenburg2016inferring, seeger2021chatbots}, i.e. supporting users in achieving their goals. For instance, previous research suggests that there is a relationship between CAs' conversation architecture and how users perceive these agents %.
%While various studies investigated how some elements of conversation architecture affect users' perceptions of agents~
(e.g.,~\cite{clark2019state, feine2019taxonomy, finch2020towards}).  However, there is currently a lack of systematic understanding of these relationships, i.e. what elements of conversation architecture are impactful for users' perceptions and how. Specifically, while various studies investigated how some elements of conversation architecture affect aspects of users' perceptions or discussed the speech specifics of agents as part of a broader analysis of the state of research \cite{clark2019state,feine2019taxonomy}, there is no comprehensive analysis of %research specifically for 
conversation architecture elements that are relevant to user perceptions. Yet, this systematic understanding is essential for guiding the design of conversation architecture, which would allow us to both predict and develop users' appropriate perceptions of agents.

@@

%These research gaps related to the paucity of knowledge on synthesizing the effect of speech specifics on perceptions of CAs as well as the lack of systematic understanding on the elements of conversation architecture and perceptions of agents 
To address this gap, %in this paper, we
%we aim to understand how to design conversation architecture in order to orchestrate perceptions of conversational agents.
%First, we 
we explore the following research questions in this paper: \textbf{(RQ1) what aspects of users' perceptions of agents have been explored in relation to the effects of conversation architecture elements?}, \textbf{(RQ2) what elements of conversation architecture are relevant to users' perceptions of agents?}, and \textbf{(RQ3) what is known about the specifics of these relationships between conversation architecture elements and the perceptions of CAs?}
%While there are various elements of conversation architecture used in a CA's utterances, we don't have a systematic understanding of them. This leads to our second research question: \textbf{(RQ2) what elements of conversation architecture are relevant to users' perceptions of agents?} Lastly, we want to understand the effect of conversation architecture on perceptions of agent, specifically: \textbf{(RQ3) what are the relationships between conversation architecture elements and the perceptions of CAs?}
To answer these questions, we performed a comprehensive review of the existing literature published from 2010 to 2022 and synthesized the effect of conversation architecture elements on the perceptions of CAs based on the final dataset of 57 relevant papers (\autoref{fig:heatmap_identified}). Based on this review, we grouped the perceptions of agents that were shown to be affected by conversation architecture into four categories (interaction, ability, sociability and humanness) and eleven aspects (\autoref{tbl:perceptions}), and grouped the speech variations related to perceptions of agents into four categories (dialog strategy, content affectiveness, content style and speech format) and eleven elements (\autoref{tbl:conversation_architecture}). 
We found that perceptions of interaction with CAs (usability, engagement, satisfaction) have been most explored in relation to speech variations, while the perceptions of CAs' ability (intelligence, competence, credibility) have been explored the least.
%the least number of explored and identified relationships. 
Our results also show that the conversation architecture elements of response delay and agent-initiated dialog related to perceptions of CAs are under-explored in literature. 
Thus, our work contributes to the CUI community by presenting a framework that synthesizes the findings on the effect of conversation architecture on perceptions of conversational agents while also identifying and clustering perception aspects and conversation architecture elements. %, providing clear and consistent terms that can be used in future research. Together, these contributions move us one step closer to designing CAs that effectively interact with users.

### 2023-02-27 sections/2-related_work.tex
submitted version to CUI 2023

\section{Related Work}
%Communication with conversational agents 
Similar to human-human communication, human communication with conversational agents relies on natural languages for interaction. However, while it is tempting to leverage components from human-human conversation in designing interactions with conversational agents, studies have demonstrated that there are differences between human-human communications and human-agent communications \cite{clark2019makes, hill2015real}. Currently, CAs are considered as a user-controlled tool rather than a social companion, with a focus on the utilitarian aspects of the conversation \cite{clark2019makes}\cmt{[1]}. When conversing with agents, users tend to use shorter messages with limited vocabulary, adapting to the style of the agent \cite{hill2015real}. Also, some studies found that small talk and humour are considered unnecessary or even inauthentic in conversations with CAs, while these elements are important in scaffolding communications with human partners \cite{clark2019makes, doyle2019mapping}\cmt{[1],[2]}. There are also machine-like traits that users preferred in a CA, such as the ability to interact through multimodal media, and a machine's perceived ability to be objective and non-judgemental \cite{doyle2019mapping, kim2022understanding}\cmt{[2],[3]}. Given human-agent communications are different from human-human communications, it is important to look specifically at the design of user experiences for conversational agents.

@@

In recent years, several papers have synthesized the current state of research for conversational agents for specific agent types (e.g. text-based, voice-based, polyadic) highlighting major trends, topics, methods and evaluating metrics \cite{clark2019state, rapp2021human, zheng2022ux}\cmt{[33][5]}. Other papers looked at specific dimensions of user experiences or domains of usage, such as \citet{van2020human} examining the effects of agents' human-like communicative behaviours on their relationships with users, and \citet{kocaballi2022design} exploring the challenges and opportunities of CAs in healthcare. There are also publications aimed to unify concepts used in the research for CA user experiences, for instance \citet{feine2019taxonomy} creating a taxonomy of social cues, and \citet{finch2020towards} analyzing evaluation protocols for dialogue systems. To the best of our knowledge, there is currently no synthesis on the speech specifics of CAs, the perceptions of agents, or the relationship between them.

>>>>

\section{Related Work}
%Communication with conversational agents 
Similar to human-human communication, users utilize natural languages when interacting with conversational agents. %human communication with conversational agents relies on natural languages for interaction. 
However, while it is tempting to leverage components from human-human conversation in designing interactions with conversational agents, studies have demonstrated that there are differences between human-human communications and human-agent communications \cite{clark2019makes, hill2015real}. Currently, CAs are considered as a user-controlled tool rather than a social companion, with a focus on the utilitarian aspects of the conversation \cite{clark2019makes}\cmt{[1]}. When conversing with agents, users tend to use shorter messages with limited vocabulary, adapting to the style of the agent \cite{hill2015real}. Also, some studies found that small talk and humour are considered unnecessary or even inauthentic in conversations with CAs, while these elements are important in scaffolding communications with human partners \cite{clark2019makes, doyle2019mapping}\cmt{[1],[2]}. There are also machine-like traits that users preferred in a CA, such as the ability to interact through multimodal media, and a machine's perceived ability to be objective and non-judgemental \cite{doyle2019mapping, kim2022understanding}\cmt{[2],[3]}. Given human-agent communications are different from human-human communications, it is important to look specifically at the design of user experiences for conversational agents.

@@

In recent years, several papers have synthesized the current state of research for particular types of conversational agents (e.g. text-based, voice-based, polyadic) highlighting major trends, topics, methods and evaluating metrics \cite{clark2019state, rapp2021human, zheng2022ux}\cmt{[33][5]}. Other papers looked at specific dimensions of user experiences or domains of usage, such as \citet{van2020human} examining the effects of agents' human-like communicative behaviours on their relationships with users, and \citet{kocaballi2022design} exploring the challenges and opportunities of CAs in healthcare. There are also publications aimed to unify concepts used in the research for CA user experiences, for instance \citet{feine2019taxonomy} creating a taxonomy of social cues, and \citet{finch2020towards} analyzing evaluation protocols for dialogue systems. To the best of our knowledge, there is currently no synthesis on the speech specifics of CAs, the perceptions of agents, or the relationship between them.

### 2023-02-27 sections/3-method.tex
submitted version to CUI 2023

The following selection criteria are applied to identify literature related to the effect of conversational architecture on the perception of agents:
\begin{itemize}
  \item The paper was published between 2010 and 2022.
  \item The paper is peer-reviewed and written in English.
  \item The paper contains the use of voice-based or text-based conversational agents.
  \item The paper contains studies on the impact of conversation architecture elements on the perceptions of agents.
  \item The paper contains impact of conversation architecture elements can be separated from other effects (e.g. embodiment).

@@

The papers reviewed (n=57) were published between May 2011 and November 2022. The majority of the papers were published in or after 2019, with a slight trend upwards over the years (see \autoref{fig:paper}). The vast majority of the papers in our corpus were published in conference proceedings (n=50), with the remainder papers published in journals. Out of the papers published in conferences, the top conferences were The ACM Conference on Human Factors in Computing Systems\footnote{https://dl.acm.org/conference/chi} (n=16), Conversational User Interfaces\footnote{https://dl.acm.org/conference/cui} (n=7), Human-Agent Interaction\footnote{https://dl.acm.org/conference/hai} (n=4), and Intelligent Virtual Agents\footnote{https://dl.acm.org/conference/iva} (n=4). 
%Out of the papers published in journals, some venues include the International Journal of Human-Computer Interaction\footnote{https://www.tandfonline.com/journals/hihc20} (n=2) and Interacting with Computers\footnote{https://academic.oup.com/iwc} (n=1).

On the modality characteristics of the conversational agents in the corpus, there is a roughly even split between the modalities, with slightly more papers exploring voice-based CAs (n=30) compared to text-based CAs (n=27).

@@

\subsubsection*{Perceptions of Agents} To understand the perceptions of agents related to conversation architecture (RQ1), assessments for the perceptions of agents were extracted from each paper. We used inductive thematic analysis to group these assessments into codes based on their meanings. For example, the code likeability of the agent contains the questionnaire item "this voice agent was likeable" used by \cite{cuadra2021my}\cmt{[67]}, Godspeed questionnaire \cite{bartneck2009measurement}\cmt{godspeed}'s set of questions on likeability used by \cite{linnemann2018can}\cmt{[15]}, and Subjective Assessment of Speech System Interface (SASSI) questionnaire \cite{hone2000towards}\cmt{sassi}'s set of questions on likeability used by \cite{chan2021kinvoices, choi2020nobody}\cmt{[74][54]}. This step resulted in 83 unique codes. These codes are then organized based on similarity of the meanings, resulting in 11 aspects of perceptions. For example, the aspect of personality traits of the agent contains measurements such as likeability, funniness, and kindness. Lastly, the 11 aspects are grouped into 4 categories: perception of interaction with agent, perception of agent's ability, perception of sociability with agent, and perception of agent's humanness (\autoref{tbl:perceptions}).

\subsubsection*{Conversation Architecture Elements}
As to the elements of conversation architecture that are relevant to the perception of agents (RQ2), we extracted the specifics of conversation architecture used in each paper. Some of the papers used multiple features in their CA design, such as the the anthropomorphic agent used in Seeger et al's study \cite{seeger2021chatbots}\cmt{[35]}. In these composite situations, we broke them down into their individual features. For example, the anthropomorphic agent \cite{seeger2021chatbots}\cmt{[35]} was separated into emotional expressions, is-typing indicator, emoticons, and response delay and captured as codes. 58 unique codes were created as part of this process, capturing features like sentiment-adaptive responses \cite{diederich2019emulating}\cmt{[25]}, lexical alignment \cite{spillner2021talk}\cmt{[18]}, and typos \cite{westerman2019believe}\cmt{[9]}. We then organized these codes into elements based on similarity, resulting in 12 elements of conversation architecture. For example, the element of disfluency contains fillers \cite{jeong2019exploring, wester2015artificial}\cmt{[10][14]}, interjections \cite{ceha2022expressive, hu2021enhancing}\cmt{[77][56]}, repetitions \cite{yang2021effect}\cmt{[72]} and typos \cite{westerman2019believe}\cmt{[9]}. These elements are grouped into 4 categories: dialog strategy, content affectiveness, content style, and speech format (\autoref{tbl:conversation_architecture}).

@@

To study the effect of conversation architecture on users' perceptions (RQ3), we extracted the connections explored in each paper, using the perception aspects and conversation architecture elements developed in the previous data analysis steps. Each connection's effect was also recorded based on whether an association was found in the study and the nature of that association. Out of our review corpus, 265 connections between perceptions of agents and conversation architecture were explored in literature. To analyze the specifics of conversation architecture that affect the perceptions of CAs, 72 connections that did not result in observed relationships were discarded, resulting in 193 relationships for analysis. For example, the connection between the matching style of a CA and user satisfaction was removed because Hoegen et al. \cite{hoegen2019end}\cmt{[31]} did not find a significant difference between the style matching agent and the non-style matching agent for overall interaction satisfaction. Out of the 193 relationships, 10 were relationships based on agents modifying multiple architecture elements (e.g. \cite{seeger2021chatbots, volkel2021manipulating}\cmt{[35][68]}). They were not included in the final framework as we could not attribute the perceptions of agents to the underlying core elements. Overall, 183 relationships between individual architecture elements and perceptions of agents are incorporated into our synthesized framework, visualized as a heatmap to demonstrate their relationships with each other (\autoref{fig:heatmap_identified}).

>>>>

The following selection criteria are applied to identify literature related to the effect of conversational architecture on the perception of agents:
\begin{itemize}
  \item The paper was published between 2010 and 2022.
  \item The paper is peer-reviewed and written in English.
  \item The paper contains the use of voice-based or text-based conversational agents.
  \item The paper contains studies on the effect of conversation architecture elements on the perceptions of agents.
  \item The paper contains effects of conversation architecture elements can be separated from other effects (e.g. embodiment).

@@

The papers reviewed (n=57) were published between May 2011 and November 2022. Most of the papers were published in or after 2019, with a slight trend upwards over the years (see \autoref{fig:paper}). The vast majority of the papers in our corpus were published in conference proceedings (n=50), while the remaining papers were published in journals. Out of the papers published in conferences, the top conferences were The ACM Conference on Human Factors in Computing Systems\footnote{https://dl.acm.org/conference/chi} (n=16), Conversational User Interfaces\footnote{https://dl.acm.org/conference/cui} (n=7), Human-Agent Interaction\footnote{https://dl.acm.org/conference/hai} (n=4), and Intelligent Virtual Agents\footnote{https://dl.acm.org/conference/iva} (n=4). 
%Out of the papers published in journals, some venues include the International Journal of Human-Computer Interaction\footnote{https://www.tandfonline.com/journals/hihc20} (n=2) and Interacting with Computers\footnote{https://academic.oup.com/iwc} (n=1).

On the modality characteristics of the conversational agents in our corpus, there is a roughly even split between voice and text modalities, with slightly more papers exploring voice-based CAs (n=30) compared to text-based CAs (n=27).

@@

\subsubsection*{Perceptions of Agents} To understand the perceptions of agents related to conversation architecture (RQ1), assessments for the perceptions of agents were extracted from each paper. We used inductive thematic analysis to group these assessments into codes based on their meanings. For example, likeability of the agent contains Godspeed questionnaire \cite{bartneck2009measurement}\cmt{godspeed}'s set of questions on likeability used by \cite{linnemann2018can}\cmt{[15]}, Subjective Assessment of Speech System Interface (SASSI) questionnaire \cite{hone2000towards}\cmt{sassi}'s set of questions on likeability used by \cite{chan2021kinvoices, choi2020nobody}\cmt{[74][54]} and questionnaire items like ``\textit{this voice agent was likeable}'' used by \cite{cuadra2021my}\cmt{[67]}. This step resulted in 83 unique codes. These codes are then organized based on similarity of their meanings, resulting in 11 aspects of perceptions. For instance, the aspect of personality traits of the agent contains codes such as likeability, funniness, and kindness. Lastly, the 11 aspects are grouped into 4 categories: perception of interaction with agent, perception of agent's ability, perception of sociability with agent, and perception of agent's humanness (\autoref{tbl:perceptions}).

\subsubsection*{Conversation Architecture Elements}
As for the elements of conversation architecture that are relevant to the perception of agents (RQ2), we extracted the speech variation features used in each paper. Some of the papers used multiple features in their CA design, such as the anthropomorphic agent used in Seeger et al's study \cite{seeger2021chatbots}\cmt{[35]}. In these composite situations, we broke the CA design down into its individual features. For example, the design of the anthropomorphic agent \cite{seeger2021chatbots}\cmt{[35]} was divided into features of emotional expressions, is-typing indicator, emoticons, and response delay and captured as separate codes. 58 unique codes were created as part of this process, capturing features like sentiment-adaptive responses \cite{diederich2019emulating}\cmt{[25]}, lexical alignment \cite{spillner2021talk}\cmt{[18]}, and typos \cite{westerman2019believe}\cmt{[9]}. We then organized these codes into elements based on similarity, resulting in 11 elements of conversation architecture. For example, the element of disfluency contains fillers \cite{jeong2019exploring, wester2015artificial}\cmt{[10][14]}, interjections \cite{ceha2022expressive, hu2021enhancing}\cmt{[77][56]}, repetitions \cite{yang2021effect}\cmt{[72]} and typos \cite{westerman2019believe}\cmt{[9]}. These elements are then grouped into 4 categories: dialog strategy, content affectiveness, content style, and speech format (\autoref{tbl:conversation_architecture}).

@@

To study the effect of conversation architecture on users' perceptions (RQ3), we extracted the connections explored in each paper, using the perception aspects and conversation architecture elements developed in the previous data analysis steps. Each connection's effect was also recorded based on whether an association was found in the study and the nature of that association. Out of our review corpus, 265 connections between perceptions of agents and conversation architecture were explored in literature. To analyze the specifics of conversation architecture that affect the perceptions of CAs, 72 connections that did not result in observed relationships were discarded, resulting in 193 relationships for analysis. For example, the connection between the matching style of a CA and user satisfaction was removed because Hoegen et al. \cite{hoegen2019end}\cmt{[31]} did not find a significant difference between the style matching agent and the non-style matching agent for overall interaction satisfaction. Out of the 193 relationships, 10 were based on comparisons between CAs with multiple speech element variations (e.g. \cite{seeger2021chatbots, volkel2021manipulating}\cmt{[35][68]}). They were not included in the final framework as we could not attribute the perceptions of agents to the individual elements. Overall, 183 relationships between individual conversation architecture elements and perceptions of agents are incorporated into our synthesized framework, visualized as a heatmap to demonstrate their relationships with each other (\autoref{fig:heatmap_identified}).

\subsection{Limitations}

Our systematic literature review was limited to papers published in the ACM Digital Library between 2010 and 2022. We may have missed literature published outside this time period, as well as in other libraries. Also, most of the studies in our corpus are based on lab experiments using short interactions with users. The generalizability of these findings needs to be verified through longer-term engagements with conversational agents deployed in real-world situations. Lastly, research without empirical studies on the perception of agents is excluded based on our selection criteria. We may have missed some influential findings in these papers. %To better understand the relationships between conversation architecture elements and the perceptions of agents, we plan to conduct future reviews in literature and user studies to understand their real-world impacts.
